### 2023-12-15 sections/2-methods.tex
changes to methods

\begin{equation}
    \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) }
    \label{eq:methods:lfd}
\end{equation}

@@

Najib We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance of any point in the cluster to the \emph{center}.
We estimate \emph{local fractal dimension} at the scale of the cluster radius and half the cluster radius.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
% Najib: Add an equation for local fractal dimension with the simplification from the previous sentence.

% Najib: Rewrite this paragraph. Old paper defined it as ... but we define it for the hierarchical clustering scheme as ...
We define the \emph{metric entropy} $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ under a hierarchical clustering scheme as a refinement of~\cite{yu2015entropy}, where metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data.
Here, we use a hierarchical, divisive clustering approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split.
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of leaf clusters.

@@

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:sharding-and-auto-tuning}
% Najib: Remove Sharding from this subsection. We do not show sharding in any of our benchmarks. All benchmarks are done with datasets that fit in memory (on giant machines, but still).

In this work, we also introduce a method of \emph{sharding} for datasets that cannot fit in memory.
This approach also leverages the fact that $\rho$-nearest neighbors search at the radius of the $k$-th farthest neighbor is typically much faster than $k$-nearest neighbors search.
In addition to sharding, we perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to be used with a given dataset.

With this approach, we begin by taking a random sample of no more than 10\% of the data.
This sample is considered the first shard.
The exact proportion of the dataset in this random sample depends on the dimensionality of the dataset and a user-specified memory limit. 
We then split the rest of the dataset into more shards based on the user-defined memory limit.
For each shard, we build a cluster tree using Algorithm~\ref{alg:methods:clustering:partition}.
From the first shard, we consider all clusters at some user-specified depth in the tree, with a default depth of 10.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}.
We select the algorithm which had the lowest total search time over all the queries.

Sharded search then proceeds as follows.
We use previously determined $k$-NN algorithm on the first shard. 
We use the distance to the $k$-th farthest neighbor to perform $\rho$-NN search on each subsequent shard. 
After searching each shard, we keep only the $k$-nearest neighbors so far, potentially decreasing the distance to the $k$-th farthest neighbor for search on the next shard.

Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.
Note also that the size of the shards is not necessarily consistent;
it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th neighbor), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the optimal ratio between the size of the first shard and the sizes of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work.
For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but, since the entire dataset fits in memory, there is only one shard.

@@

% Najib: For this figure, I want the red, green and blue lines/braces to be thinner. Perhaps 1/2 to 2/3 as thick.
% Najib: Also, add a caption.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.png}
    \caption{}
    \label{fig:methods:deltas}
\end{figure}

@@

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which could contain points in the query ball. 
% Najib: This is laid out in Algorithm 2. We should reference that algorithm here.

% Najib: Merge Figures 2 and 3. There's too much redundancy between them.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-1.pdf}
    \caption{
        Triangle formed by query $q$, and poles $\ell$ and $r$. Query ball of radius $\rho$ centered at $q$. 
    }
    \label{fig:methods:overlapping-children-1}
\end{figure}

To determine whether both children can contain points in the query ball, we consider the following geometry. 
Let $\ell$ and $r$ denote the cluster's left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}), $q$ denote the query, and $\rho$ denote the search radius, as shown in Figure~\ref{fig:methods:overlapping-children-1}.
Without loss of generality, we assume that $\overline{qr \vphantom \ell} \leq \overline{q\ell}$ (i.e., the query is closer to the right pole than to the left pole).

@@

@@

% Najib: I don't like the double vertical bars for norm-notion. I think it's too messy. We should just overload the notation of the single horizontal bar across the top.
% Najib: Remove the equation numbers from this paragraph. They're not necessary.
We can do a quick check using values computed during partitioning (see Section~\ref{subsubsec:methods:cluster-partitioning}) to determine whether $d \leq \rho$. 
We note that $d = ||\overline{mq'}|| = ||\overline{mr \vphantom {q'}}||-||\overline{q'r}||$. Since $m$ is the midpoint of $\overline{\ell r}$, we have that $||\overline{mr}|| = \tfrac{||\overline{\ell r}||}{2}$, so \begin{equation} d = \tfrac{||\overline{\ell r}||}{2} - ||\overline{q'r}||.
\label{eq:methods:d-expression}
\end{equation}

It remains to find $||\overline{q'r}||$. To do this, we use the Law of Cosines with $\triangle \ell rq$. Let $\theta$ denote $\angle lrq$, as shown in Figure~\ref{fig:methods:overlapping-children-3}. By the Law of Cosines, we have that 
\begin{equation} \text{cos}(\theta) = \tfrac{||\overline{qr\vphantom \ell}||^2 + ||\overline{\ell r}||^2 - ||\overline{q\ell}||^2}{2||\overline{\ell r}||||\overline{qr \vphantom \ell}||}.\end{equation}
On the other hand, since $\triangle q'qr$ is a right triangle,

\begin{equation} \text{cos}(\theta) = \tfrac{||\overline{q'r}||}{||\overline{qr \vphantom \ell}||}.\end{equation}

Equating the two expressions for $\text{cos}(\theta)$ and solving for $||\overline{q'r}||$, we have \begin{equation} ||\overline{q'r}|| = \tfrac{||\overline{qr\vphantom \ell}||^2 + ||\overline{\ell r}||^2 - ||\overline{q\ell}||^2}{2 ||\overline{\ell r}||}. \end{equation}

Substituting this back in equation~\ref{eq:methods:d-expression} gives 
\begin{equation} d = \tfrac{||\overline{q\ell}||^2 - ||\overline{qr \vphantom \ell}||^2}{2||\overline{lr}||}.\end{equation}

Then to check if $d \leq \rho$, it suffices to verify that \begin{equation}(||\overline{q\ell}|| + ||\overline{qr \vphantom \ell}||)(||\overline{q\ell}||- ||\overline{qr \vphantom \ell}||) \leq 2||\overline{\ell r}||\rho. \end{equation}

% Najib: Move the cluster.isLeaf check to the beginning of the algorithm because it's the base case.
\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE{$\rho$-NN($cluster.left, query, r$)}
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

>>>>

\begin{equation}
    \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) }
    \label{eq:methods:lfd_original}
\end{equation}

@@

Najib We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance of any point in the cluster to the \emph{center}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
We estimate local fractal dimension at the scale of the cluster radius and half the cluster radius. 
This allows us to rewrite Equation \ref{eq:methods:lfd_original} as simply: 
\begin{equation} 
\log_2 \frac{|B_X(q, r)}{|B_X(q, r/2)|}
\end{equation}

% Najib: Rewrite this paragraph. Old paper defined it as ... but we define it for the hierarchical clustering scheme as ... DONE. (let me know if this okay)

In~\cite{yu2015entropy}, metric entropy for a given cluster radius $r_c$ was defined as the number of clusters of radius $r_c$ needed to cover all data. In this paper, we define the metric entropy $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree.

% Here, we use a hierarchical, divisive clustering approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split.
% Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of leaf clusters.

@@

\subsection {Auto-Tuning}
\label{subsec:methods:auto-tuning}

% In this work, we also introduce a method of \emph{sharding} for datasets that cannot fit in memory.
% This approach also leverages the fact that $\rho$-nearest neighbors search at the radius of the $k$-th farthest neighbor is typically much faster than $k$-nearest neighbors search.

We perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
With this approach, we begin by taking a random sample of no more than 10\% of the data.
We build a cluster tree with this sample using Algorithm~\ref{alg:methods:clustering:partition}, and select all clusters at some user-specified depth in the tree, with a default depth of 10.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}.
We select the algorithm which had the lowest total search time over all the queries. 
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

% Note also that the size of the shards is not necessarily consistent;
% it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
% However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th neighbor), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
% Finding the optimal ratio between the size of the first shard and the sizes of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work.
% For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but, since the entire dataset fits in memory, there is only one shard.

@@

% Najib: For this figure, I want the red, green and blue lines/braces to be thinner. Perhaps 1/2 to 2/3 as thick.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.png}
    \caption{$\delta$, $\delta_{\max}$, and $\delta_{\min}$ defined for cluster $C$ and query $q$.}
    \label{fig:methods:deltas}
\end{figure}

@@

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which could contain points in the query ball. The full algorithm is described in Algorithm~\ref{alg:methods:rho-nn-search}.

% Najib: Merge Figures 2 and 3. There's too much redundancy between them.
% \begin{figure}[ht!]
%     \centering
%     \includegraphics[scale=0.5]{images/geometry/overlapping-children-1.pdf}
%     \caption{
%         Triangle formed by query $q$, and poles $\ell$ and $r$. Query ball of radius $\rho$ centered at $q$. 
%     }
%     \label{fig:methods:overlapping-children-1}
% \end{figure}

@@

To determine whether both children can contain points in the query ball, we consider the following geometry. 
Let $\ell$ and $r$ denote the cluster's left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}), $q$ denote the query, and $\rho$ denote the search radius, as shown in Figure~\ref{fig:methods:overlapping-children-3}.
Without loss of generality, we assume that $\overline{qr \vphantom \ell} \leq \overline{q\ell}$ (i.e., the query is closer to the right pole than to the left pole).

@@

We can do a quick check using values computed during partitioning (see Section~\ref{subsubsec:methods:cluster-partitioning}) to determine whether $d \leq \rho$. 
In the caculations that follow, we overload notation a bit and use $\overline{xy}$ to mean the length of the line segment between $x$ and $y$. We note that $d = \overline{mq'} = \overline{mr \vphantom {q'}}-\overline{q'r}$. Since $m$ is the midpoint of $\overline{\ell r}$, we have that $\overline{mr} = \tfrac{\overline{\ell r}}{2}$, so \begin{equation*} d = \tfrac{\overline{\ell r}}{2} - \overline{q'r}.
\label{eq:methods:d-expression}
\end{equation*}

It remains to find $\overline{q'r}$. To do this, we use the Law of Cosines with $\triangle \ell rq$. Let $\theta$ denote $\angle lrq$, as shown in Figure~\ref{fig:methods:overlapping-children-3}. By the Law of Cosines, we have that 
\begin{equation} \text{cos}(\theta) = \tfrac{\overline{qr\vphantom \ell}^2 + \overline{\ell r}^2 - \overline{q\ell}^2}{2(\overline{\ell r})(\overline{qr \vphantom \ell})}.\end{equation}
On the other hand, since $\triangle q'qr$ is a right triangle,

\begin{equation} \text{cos}(\theta) = \tfrac{\overline{q'r}}{\overline{qr \vphantom \ell}}.\end{equation}

Equating the two expressions for $\text{cos}(\theta)$ and solving for $\overline{q'r}$, we have \begin{equation} \overline{q'r} = \tfrac{\overline{qr\vphantom \ell}^2 + \overline{\ell r}^2 - \overline{q\ell}^2}{2 \overline{\ell r}}. \end{equation}

Substituting this back in equation~\ref{eq:methods:d-expression} gives 
\begin{equation} d = \tfrac{\overline{q\ell}^2 - \overline{qr \vphantom \ell}^2}{2\overline{lr}}.\end{equation}

Then to check if $d \leq \rho$, it suffices to verify that \begin{equation}(\overline{q\ell} + \overline{qr \vphantom \ell})(\overline{q\ell}- \overline{qr \vphantom \ell}) \leq 2\overline{\ell r}\rho. \end{equation}

% Najib: move is.Leaf check to the beginning of the algorithm because it's the base base case. DONE (confirm this is correct)

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
            \STATE Return $results$
        \ENDIF
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE{$\rho$-NN($cluster.left, query, r$)}
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF

    \end{algorithmic}
    \label{alg:methods:rho-nn-search}
\end{algorithm}
